---
title: "sim_filter_arbitrary"
author: "Jonsen et al"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## install aniMotum dev branch
```{r, eval=FALSE}
remotes::install_github("ianjonsen/aniMotum@dev", dependencies = TRUE)
```

```{r load pkgs, message=FALSE}
require(tidyverse, quietly = TRUE)
require(aniMotum, quietly = TRUE) 
```

## Fit `crw` SSM with 12 h time.step
```{r fit SSM}
fit <- fit_ssm(sese2, model = "crw", time.step = 12,
               control = ssm_control(verbose = 0))
```

## Simulate 200 tracks from the 2nd fitted track, ensuring they return to the start & plot
```{r simulate}
trs <- sim_fit(fit[2,], what = "p", reps = 200, cpf = TRUE)

plot(trs)
```

## Filter simulated tracks using similarity flag = 1, keep top 30 % & plot
```{r filter w flag 1}
trs.f1 <- sim_filter(trs, keep = 0.3, flag = 1)

plot(trs.f1)
```

## Filter simulated tracks using mean lon-lat as similarity metrix, keep top 30 % & plot
```{r filter w lonlat}
trs.fll <- sim_filter(trs, keep = 0.3, var = c("lon","lat"), FUN = "mean")

plot(trs.fll)
```

#### Comparison of mean lon-lat simulated track locations to the fitted track appears to give a slightly better match than similarity flag 1. 

## Use path re-routing to move simulated tracks off of land
```{r rr}
trs.fllr <- route_path(trs.fll, centroids = TRUE)

plot(trs.fllr)
```

## Use Chl raster & extract values at track locations
```{r arbitrary var}
## read SpatRaster
chl <- terra::rast("chl.grd")

terra::plot(chl)

trs.df <- unnest(trs.fllr, cols = c(sims))

## extract chl values at track locations
trs.df <- trs.df |> 
  mutate(terra::extract(chl, cbind(lon, lat))) |>
  rename(chl = chl_summer_climatology)

## convert back to nested tibble
trs.n <- nest(trs.df, sims = c(rep, date, lon, lat, x, y, chl))

## append new nested tibble with correct aniMotum classes so sim_filter works
class(trs.n) <- append(class(trs.fllr)[1:2], class(trs.n))
```

## Run sim_filter to select top 15% of tracks with chl most similar to the SSM-estimated track
```{r filter chl}
## try filtering on mean chl
trs.chl.mn <- sim_filter(trs.n, keep = 0.15, var = "chl", FUN = "mean", na.rm = TRUE)
```

## Visualise
```{r chl-track plots}
plot(trs.chl.mn)

## all tracks
tmp.a <- unnest(trs.n, cols = c(sims))

## filtered tracks
tmp.f <- unnest(trs.chl.mn, cols = c(sims))

## Kernel densities of all Chl values
d.all <- density(tmp.a$chl, na.rm = TRUE)

## densities of Chl values from estimated track
d.e <- density(subset(tmp.f, rep == 0)$chl, na.rm = TRUE)

## densities of Chl values from filtered random tracks
d.s <- density(subset(tmp.f, rep > 0)$chl, na.rm = TRUE)

## plot density of all chl values
plot(d.all, ylim = c(0, 1.75), lwd = 2)

## add density of Chl from estimated track
lines(d.e, col = "red", lwd = 2)

## add density of Chl from filtered random tracks
lines(d.s, col = "blue", lwd = 2)

```

