---
title: "Mapping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
require(foieGras)
```

## `map()`
`foieGras` provides a flexible `map` function for mapping SSM-estimated tracks along with options for including their confidence ellipses, observed locations, track lines, and behavioural estimates. Default choices are made based on the model fit object being mapped but these can be overridden via the `aes` argument and the `aes_lst()` function. 

#### Map an SSM fit to data, using default settings
```{r map 1, fig.height=4, fig.width=7, warning=FALSE, message=FALSE}
fit <- fit_ssm(sese, 
               model = "mp", 
               time.step = 24, 
               control = ssm_control(verbose = 0))

map(fit, what = "predicted")
```

Here, SSM-predicted locations are coloured by the estimated move persistence index ($\gamma_t$). To aid visualisation of high and low move persistence regions along each track, the $\gamma_t$ estimates are normalised separately for each track to span the interval 0,1. This can be turned off with `normalise = FALSE`, or applied to the group of tracks with `group = TRUE`.

By default, `map()` will try to use `rnaturalearth::ne_countries(scale = 10)` high-resolution land polygon data if the `rnaturalearthhires` data package is installed. If not, the default will be dropped to the coarser `scale=50`. The `rnaturalearthhires` package can be installed via: `install.packages("rnaturalearthhires", repos = "https://ropensci.r-universe.dev")` or `remotes::install_github("ropensci/rnaturalearthhires")`. Alternatively, if the `ggpstial` and `rosm` packages are installed, users may specify any of the available tile map layers returned by `rosm::osm.types()` via the `map_type` argument. When specifying a `map_type`, `map` will take any additional arguments to `ggspatial::annotation_map_tile()` for finer control; `zoom` and `progress` are used here.

```{r 1b, fig.height=4, fig.width=7, message=FALSE, warning=FALSE}
map(fit, what = "predicted", map_type = "cartodark", zoom = 4, progress = "none")
```

#### Control track components
We can add and remove certain track components by using the `aes` argument and `aes_lst()` function. Here, the observed locations (orange triangles) and track lines have been turned on and the move persistence index has been turned off.
```{r map 2, warning=FALSE, message=FALSE, fig.width=7, fig.height=4}
my.aes <- aes_lst(obs=TRUE, line=TRUE, mp=FALSE)

map(fit, what = "p", aes = my.aes)
```

Note that when turning off the move persistence index, tracks are automatically given unique colours. This can be turned off by using `by.id = FALSE`. 

Additional control over mapping aesthetics can be gained by viewing and editing the `my.aes$df` data.frame created by `aes_lst()`. We can subset components and assign new values as needed, here we change the observation colour and shape from orange triangles to firebrick diamonds. We also set `by.id = FALSE` so all tracks are displayed in a unique colour (default is "dodgerblue").
```{r map 3, warning=FALSE, message=FALSE, fig.width=7, fig.height=4}
my.aes$df

my.aes$df$col[4] <- "firebrick"
my.aes$df$shape[4] <- 18

map(fit, 
    what = "p", 
    aes = my.aes, 
    by.id = FALSE)
```

Alternatively, we can colour estimated locations by their date to better see how tracks evolve over time. 
```{r map 4, warning=FALSE, message=FALSE, fig.width=7, fig.height=4}
map(fit, 
    what = "p", 
    aes = aes_lst(mp=FALSE, conf=FALSE), 
    by.date = TRUE, 
    by.id=FALSE)
```

If you don't like the default colour palettes, these can be changed by modifying the default `aes` object. Here, the `hcl.colors()` function is used to specify a new colour palette for displaying date-along-track.
```{r map 5, warning=FALSE, message=FALSE, fig.width=7, fig.height=4}
map(fit, 
    what = "p", 
    aes = aes_lst(mp=FALSE, 
                  conf=FALSE, 
                  date_pal = hcl.colors(n=100, "Plasma")), 
    by.date = TRUE, by.id=FALSE)
```

And similarly for displaying move persistence.
```{r map 6, warning=FALSE, message=FALSE, fig.width=7, fig.height=4}
map(fit, 
    what = "p", 
    aes = aes_lst(mp_pal = hcl.colors(n=100, "RdBu")))
```

#### Change projection
Often choosing an appropriate map projection can provide a more meaningful view of the tracks, especially for animals that make very broad-scale movements, such as the southern elephant seals displayed here. This can be done easily by providing a valid `proj4string` for the Coordinate Reference System. Here, we project the tracks onto a polar stereographic grid with longitude centered on 68^o^ East, Iles Kerguelen.
```{r map 7, warning=FALSE, message=FALSE, fig.width=7, fig.height=4}
map(fit, 
    what = "p", 
    aes = aes_lst(mp_pal = hcl.colors(n=100, "RdBu")), 
    crs = "+proj=stere +lon_0=68 +datum=WGS84 +units=km")
```





