---
title: "model_inspection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{model_inspection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(foieGras)
```

Just like any other statistical model fitting exercise, inspecting SSM fits to tracking data is an essential part of any analysis. SSM fits can be visualized quickly using the generic `plot` function on model fit objects:

```{r visualise SSM fit, fig.width=7, fig.height=5}

fit.rw <- fit_ssm(ellie, model = "rw", time.step = 12, control = ssm_control(verbose = 0))

plot(fit.rw, what = "fitted")

plot(fit.rw, what = "predicted")
```

Both fitted and predicted locations (red) can be plotted as 1-D time-series on top of the observations (blue) to visually detect any lack of fit. Observations that failed to pass the `prefilter` stage prior to SSM fitting (gold x's) can be included (default) or omitted with the `outlier` argument. A rug plot along the x-axis aids detection of data gaps in the time-series. Note, in second plot, the larger standard errors for predicted locations through small data gaps.

The SSM fits can also be visualised as 2-D tracks via the `type` argument:
```{r visualise as track, fig.width=7, fig.height=6}

plot(fit.rw, "p", type = 2, alpha = 0.02)
```

This option can provides an intuitive view of the estimated track (red) through the observations (blue), along with a 2-dimensional representation of the location uncertainty (red ellipses). Here, the alpha (error ellipse opacity), argument is set to a small value so that the many overlapping error ellipses don't overwhelm visualization of the location point estimates.

Residual plots are important for validating models, but classical Pearson residuals, for example, are not appropriate for state-space models. Instead, a prediction (one-step-ahead) residual, provides a useful, albeit computationally demanding, alternative. In `foieGras`, prediction residuals from state-space model fits are calculated using the `osar` function and can be visualized as time-series plots, Q-Q plots, or autocorrelation functions:
```{r osar plots, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
# use patchwork package to arrange plot.osar options
require(patchwork)
# calculate & plot residuals
res.rw <- osar(fit.rw)

(plot(res.rw, type = "ts") | plot(res.rw, type = "qq")) / 
  (plot(res.rw, type = "acf") | plot_spacer())
```

Here, the 3 residual plots highlight a poor fit of the `rw` SSM to the x-coordinate. The time-series residual plot implies a bias in the `rw` SSM fit over the latter half of the track, the Q-Q plot highlights a distinct departure from normality in the x residuals, and the x residuals are positively autocorrelated.

Fitting the `crw` SSM to the same data, we can see the prediction residual plots imply a less biased fit with approximately normal residuals that have little autocorrelation in both coordinates. Additionally, the `crw` model fit has a lower AIC~c~ than the `rw`. Although note that AIC statistics can be misleading for time-series models and should not be used as the sole criterion for preferring one model fit over another. Here, at least, the AIC~c~ values are in agreement with the prediction residual diagnostics.
```{r fit crw and plot osar, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}

fit.crw <- fit_ssm(ellie, model = "crw", time.step = 12, control = ssm_control(verbose = 0))

res.crw <- osar(fit.crw)

(plot(res.crw, type = "ts") | plot(res.crw, type = "qq")) / 
  (plot(res.crw, type = "acf") | plot_spacer())

fit.rw$ssm[[1]]$AICc
fit.crw$ssm[[1]]$AICc
```

As calculation of prediction residuals can be computationally demanding, typically requiring more time than fitting the model, especially for multiple individual tracks, the `osar` function is automatically implemented in parallel when calculating residuals for more than 2 tracks. 