---
title: "Map projections and your results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Map projections and your results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
require(aniMotum)
```

## The short version

`aniMotum` estimates locations and movement parameters in map coordinates, not
in longitude and latitude. That means the map you use is part of the model. If
your animals range across a wide band of latitude, the default map will stretch
their movements by different amounts in different places, and the parameter
estimates you get back will partly describe the map rather than the animal.

For tracks that stay within a narrow band of latitude, this makes almost no
difference and you can ignore it. For tracks that cross a wide band - polar and
sub-polar animals especially - it can change the answers substantially.

## What `aniMotum` does by default

Tracking data usually arrive as longitude and latitude. Those are angles, not
distances, so before fitting a model `aniMotum` converts them to x and y
coordinates measured in kilometres. Unless you tell it otherwise, it uses a
Mercator grid to do that.

Mercator is a sensible general-purpose default. It is the map most people
picture, it never distorts an animal's direction of travel at a point, and it
works anywhere in the world without you having to choose anything.

It has one property that matters here: it stretches distances by more and more
as you move away from the equator. The amount of stretch is `1 / cos(latitude)`.

```{r stretch}
lat <- c(0, 30, 45, 50, 60, 70, 80)
data.frame(latitude = lat,
           stretch  = round(1 / cos(lat * pi / 180), 2))
```

So a 10 km move at 70 degrees south appears on a Mercator grid as roughly a
29 km move, while the same 10 km move at 30 degrees south appears as about
11.5 km. The animal did the same thing in both cases. The map did not.

## Why that reaches the parameter estimates

The models estimate how far an animal typically moves in a given time. In the
`mp` and `rw` models that quantity is `sigma`; in the `crw` model it appears as
the diffusion coefficient `D`. Either way it is measured in map units. If the
map stretches distances, the estimate is stretched with them.

Two things follow.

**An animal that spends its time at high latitude looks faster than one at
lower latitude, even if they move identically.** The difference is entirely the
map.

**A single animal that travels across a wide band of latitude has its own
movements stretched by different amounts at different points of its own
track.** The model fits one value of `sigma` to the whole track, so what you
get is an average across all that stretching.

The measurement error parameters are affected in the same way. `tau` scales the
assumed Argos error, and that error is also expressed in map units.

## A worked example

The `sese` data are four southern elephant seals tagged at Iles Kerguelen. They
leave from roughly the same place and then travel in very different directions:
one east, one west, one almost due south, one southwest. Between them they
cover 47 to 69 degrees south.

```{r latrange}
range(abs(sese$lat))

## the average latitude each seal occupies, and the stretch that implies
ll <- tapply(abs(sese$lat), sese$id, mean)
round(1 / cos(ll * pi / 180), 2)
```

The four seals sit at stretch factors of roughly 1.6 to 2.2. So before any
model is fitted, we should expect the seals that go furthest south to look like
faster movers.

Fitting the joint move persistence model on the default Mercator grid, and then
again on an equal-area grid centred on the study region:

```{r fits, eval = FALSE}
## default: Mercator
fm <- fit_ssm(sese, vmax = 4, model = "jmp", time.step = 24)

## equal-area, centred on the deployment
sese_ea <- sf::st_transform(
  sf::st_as_sf(sese, coords = c("lon", "lat"), crs = 4326),
  "+proj=laea +lat_0=-57 +lon_0=70 +datum=WGS84 +units=km +no_defs")

fe <- fit_ssm(sese_ea, vmax = 4, model = "jmp", time.step = 24)
```

The estimates:

| parameter | Mercator | equal-area |
|---|---|---|
| `sigma_pop_x` | 3.31 | 0.71 |
| `sigma_pop_y` | 2.44 | 0.71 |
| `sigma_g` | 0.023 | 0.042 |
| `tau_x` | 0.046 | 0.087 |
| `tau_y` | 0.071 | 0.080 |
| `rho_o` | 0.51 | −0.04 |
| `sd_lsig` | 0.25 | 0 |

Three things are worth pulling out.

**The seals only appear to differ on the Mercator grid.** `sd_lsig` is how much
the seals differ from one another in movement scale. On the Mercator grid it
comes out at 0.25. On the equal-area grid it goes to zero: once the map stops
stretching the southern seals more than the northern ones, there is no
difference left to find. The apparent variation between animals was the map.

**The Mercator fit says the seals move differently east-west than north-south,
and the equal-area fit says they do not.** On the Mercator grid `sigma_pop_x`
and `sigma_pop_y` differ by about a third. On the equal-area grid they are
equal to three decimal places. Mercator stretches north-south and east-west by
the same amount at any single point, so this is not a direct distortion. It
arises because the seals do their east-west travelling and their north-south
travelling at different latitudes, so the two get stretched by different
average amounts over a whole track.

**The measurement error estimates give an independent check, and they favour
the equal-area fit.** Argos location error has no reason to be larger
east-west than north-south, and no reason for the error in one direction to be
correlated with the error in the other. On the equal-area grid that is what we
get: `tau_x` and `tau_y` are within 8% of each other, and `rho_o` is close to
zero. On the Mercator grid they differ by more than half, and `rho_o` comes out
at 0.51. The equal-area fit recovers a sensible measurement error model without
being asked to; the Mercator fit does not.

## Move persistence is affected too

It is tempting to assume that move persistence, `g`, escapes all of this. It is
a number between 0 and 1 with no units, so a change in map scale should not
touch it.

That reasoning does not hold. `g` measures how much of an animal's previous
step carries over into its next one, and a step has a direction as well as a
length. Any map that distorts shape, and any map whose stretch changes along
the track, changes the relationship between one step and the next. In the
example above `sigma_g` nearly doubled between the two fits.

So `g` values are comparable between animals in the same fit, which is what
fitting a joint model buys you, but they are not comparable across fits made on
different maps. If you report move persistence, report the projection you used.

## Does this matter for my data?

A quick check:

```{r check}
## how much does the stretch vary across your data?
lat.rng <- range(abs(sese$lat), na.rm = TRUE)
stretch <- 1 / cos(lat.rng * pi / 180)

round(stretch, 2)          # least and most stretched
round(max(stretch) / min(stretch), 2)   # how many times more one than the other
```

As a rough guide, if that number is below about 1.2 the default is fine - and
that is the threshold `projection = "auto"` uses before it changes anything.
Above about 1.5 it is worth fitting in a projection suited to your study region
and comparing. `fit_ssm()` will also warn you when a joint fit spans a wide
enough band for this to bite.

The same check applies within a single animal. A seal that migrates from 50 to
70 degrees south has this problem on its own, with no other animals involved.

## What to do about it

### Let `fit_ssm()` choose

```{r auto, eval = FALSE}
fit <- fit_ssm(mydata, vmax = 4, model = "mp", time.step = 24,
               projection = "auto")
```

`projection = "auto"` picks a projection from the geographic extent of your
data. You can see what it would choose without fitting anything:

```{r autocrs}
auto_crs(sese$lon, sese$lat)
```

The choice is made once from all the tracks together, never one per animal.
Fitting each animal in its own projection would leave their estimates in
different frames, so you could not compare them - which is usually the reason
for fitting a whole deployment.

The rules are:

- a **conformal conic** (`lcc`) for tracks within one hemisphere and below 80
  degrees, with its two standard parallels placed to keep scale error small
  across the band your animals occupy;
- **polar stereographic** (`stere`) beyond 80 degrees, or for tracks that go
  right around the pole;
- **oblique stereographic** for tracks straddling the equator, where a conic
  does not apply;
- the **existing Mercator default**, unchanged, when the stretch varies by less
  than 20 percent across your data and your animals stay below 60 degrees. For
  most temperate and tropical studies `"auto"` therefore changes nothing.

All of these keep shape undistorted - a step of a given length maps to the same
coordinate length whichever way it points. That property matters more than it
might seem, because these models are about movement *steps*, which have a
direction as well as a length. A projection that distorts shape makes the
east-west and north-south parts of the process differ for reasons that have
nothing to do with the animal, and it changes move persistence.

That last point is why an equal-area projection is not the obvious choice it
first appears. Equal-area keeps areas honest but distorts shape away from its
centre, which is exactly the wrong trade for a movement model. The equal-area
fit in the table above is much better than the Mercator one, but a conformal
projection is better still.

### Or choose one yourself

If you have a projection your study already uses, supply the data as an `sf`
object and `aniMotum` will use it unchanged:

```{r howto, eval = FALSE}
d <- sf::st_as_sf(mydata, coords = c("lon", "lat"), crs = 4326)
d <- sf::st_transform(d, "+proj=stere +lat_0=-90 +lon_0=70 +units=km")

fit <- fit_ssm(d, vmax = 4, model = "mp", time.step = 24)
```

Note the `units=km`. `aniMotum` works in kilometres internally because it helps
the models converge, and it will convert for you, but stating it avoids doubt.

**An `sf` object is always respected.** `aniMotum` only chooses a projection
for itself when the data arrive as plain longitude and latitude. If your data
have already been prepared and projected somewhere else - by `ArgosQC`, for
example - they reach the model exactly as prepared, and `projection` is
ignored. Nothing is ever projected twice.

If you are unsure, fit twice and compare. Estimates that hold steady across two
sensible projections are ones you can report with some confidence. Estimates
that move a lot are telling you something about your map.

## What this does not change

Estimated locations are barely affected. The models are estimating where the
animal was, and converting back to longitude and latitude at the end undoes
most of the distortion. If all you want from `aniMotum` is a quality-controlled
track, the default is fine.

It is the *parameters* - how fast, how variable between animals, how persistent
- that need the care, along with anything computed from them.
